---
title: "Reproducible Research: Peer Assessment 1"
author: "Mark Cohen"
output: 
  html_document:
    keep_md: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE)
```


## Required packages
This analysis makes use of the `tidyverse` packages, as well as the 'lubridate' date manipulation tools.
```{r packages}
library(tidyverse)
library(lubridate)
```


## Loading and preprocessing the data
The data is loaded directly from activity.zip into a 17568 x 3 tibble.
```{r}
activity_data <- read_csv("activity.zip")
activity_data
```
The columns are correctly formatted by `readr`'s default type interpretation.


## What is mean total number of steps taken per day?
We can first summarize the data in terms of the total number of steps recorded for each day.
```{r}
daily_steps <- activity_data %>%
    group_by(date) %>% summarize(steps = sum(steps, na.rm = TRUE))
qplot(daily_steps$steps, fill = I("blue"), bins = 30, 
      main = "Figure 1: Histogram of total steps per day", 
      xlab = "Number of steps")
```

The median number of steps recorded per day is `r median(daily_steps$steps) %>% prettyNum(format = "d", big.mark = ",")`, while the mean is `r prettyNum(mean(daily_steps$steps), format = "d", big.mark = ",")`.


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
