---
title: "Reproducible Research: Peer Assessment 1"
author: "Mark Cohen"
output: 
  html_document:
    keep_md: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE)
```


## Required packages
This analysis makes use of the `tidyverse` packages, as well as the `lubridate` date manipulation tools and `xtable` for html table formatting.
```{r packages}
library(tidyverse)
library(lubridate)
library(xtable)
```


## Loading and preprocessing the data
The data is loaded directly from activity.zip into a 17568 x 3 tibble.
```{r}
activity_data <- read_csv("activity.zip")
activity_data
```
The columns are correctly formatted by `readr`'s default type interpretation.


## What is mean total number of steps taken per day?
We can first summarize the data in terms of the total number of steps recorded for each day.
```{r}
daily_steps <- activity_data %>%
    group_by(date) %>% summarize(steps = sum(steps, na.rm = TRUE))
qplot(daily_steps$steps, fill = I("blue"), bins = 30, 
      main = "Figure 1: Histogram of total steps per day", 
      xlab = "Number of steps")
```

The median and mean of the daily steps per day are:
```{r mean_median, results="asis"}
summarize(daily_steps, Mean=mean(steps), Median=median(steps)) %>% 
    xtable(digits=c(0,2,0)) %>% print(type = "html")
```


## What is the average daily activity pattern?
By re-grouping the original data by 5-minute interval and averages the daily observations, we can see the average pattern of number of steps at different times of day.
```{r}
daily_pattern <- activity_data %>%
    group_by(interval) %>% summarize(steps = mean(steps, na.rm = TRUE))
qplot(interval, steps, data = daily_pattern, 
      geom = c("line"), 
      ylab = "Average steps", 
      xlab = "5-minute intervals", 
      main = "Figure 2: Average Daily Pattern of Steps")
```



## Imputing missing values
A substantial number of observations, `r sum(is.na(activity_data$steps))` (or `r round(mean(is.na(activity_data$steps)) * 100, 2)`% on date-interval pairs), are missing. These missing observation are clustered in eight specific dates:
```{r}
activity_data %>% group_by(date) %>% summarize("Share missing" = mean(is.na(steps))) %>% filter("Share missing" > 0)
```

As such, other observations from a given day cannot be used to impute missing values. A simple alternative strategy would be to use the average number of steps in a given interval.
```{r}
activity_imputed <- activity_data %>% group_by(interval) %>% 
    mutate(steps = if_else(is.na(steps), mean(steps, na.rm = TRUE), steps))
```

We can then recompute the daily totals of steps with the imputed data.
```{r}
daily_imputed <- activity_imputed %>%
    group_by(date) %>% summarize(steps = sum(steps, na.rm = TRUE))
qplot(daily_imputed$steps, fill = I("blue"), bins = 30, 
      main = "Figure 3: Histogram of total steps per day with missing values imputed",
      xlab = "Number of steps")
```

What we see is that the substantial number of days with 0 recorded steps reported in figure 1 above was the result of the days lacking any valid observations. This is also reflected in a higher mean (`r mean(daily_imputed$steps) %>% formatC(format = "f", digits = 2)`) and median (`r median(daily_imputed$steps) %>% formatC(format = "f", digits = 2`) daily steps, once missing values are imputed.


## Are there differences in activity patterns between weekdays and weekends?


